Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-07-18 submission/Response_to_reviewers.md

**P.10 (line 204): When discussing biological relevance, it would be useful to also mention that relative abundance may not be the relevant factor for the host. Imagine an OTU (A) that produces a metabolite with a physiological effect on obesity (or whatever host trait one is studying). Suppose the number of cells and the amount of metabolite produced remains constant, but the number of cells for some other OTU (B) goes up by an order of magnitude. The relative abundance of A in the community has gone down, but this may be irrelevant to the molecular/cellular mechanism affecting obesity, especially if B has no relationship to A or obesity (e.g., it occupies a distinct niche in the gut). The same argument probably holds for alpha diversity metrics based on relative abundances - the host likely cares about specific components of the diversity not diversity per se.**

We agree with this sentiment and included text at L258-267 to point out that there are clear examples where changing the microbiome changes BMI profile. We have expanded on this point as suggested by the reviewer below.

**P.12 (line 244): Where can the pipeline be obtained? This info is in the methods but should probably be highlighted.**

We have added a link to the GitHub repository that contains the pipeline for the analysis and this manuscript.

@@

**Reviewer #2 (Comments for the Author):

General comments:
Overall this paper is a useful meta analysis of obesity and the microbiome. I have some miscellaneous comments that are below. In addition I have one more important concern that I believe needs to be addressed.

More important concern:
Given that their analysis depends on them compiling data from multiple studies and reanalyzing that data, I believe it is necessary for these authors to make sure that all components of this analysis are reproducible. I commend the authors for their commitment to sharing code for all the work. However, I have a few concerns about the data. First, some of the data is from a personal web site. I would recommend that the authors here either repost that data to a more sustainable site (e.g., Figshare), get the original producers of the data to post it to a more sustainable site, or come up with some other solution. Again, I commend the authors for their efforts in openness and reproducibility, I am just concerned that some aspects of this paper will not be reproducible.**

We appreciate the reviewer's concern regarding the openness of the underlying data used in this study. As far as we can tell all of the sequence data is available either through the ENA, SRA, or dbGaP except for the Turnbaugh and Schubert datasets. Given the age of this dataset and the lack of access to the original sff files (we've asked previously) it is unlikely that continued requests to the Gordon lab will get these data to be made any more public. Furthermore, this is a keystone study in this research area because it was the original study to claim a relationship. The Schubert dataset is now available on the SRA (accession SRP078489). The reviewer is correct that some of the metadata is not publicly available (i.e. Wu and Zeevi). These were obtained from the original authors. We have encouraged the authors to make these data more available. Regardless of the outcome of that encouragement, we have posted the metadata for each study in the GitHub repository for this study. This is obviously a complex problem and we can only do our best to encourage our colleagues and follow best practices ourselves. Ultimately, we do not think it is necessary to censor datasets from the study because their data are not publicly available.

### 2016-07-19 submission/Sze_Obesity_mBio_2016.tex
Re-run make write.paper

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

\section{Lack of power and small effect size confounds the ability to
differentiate non-obese and obese individuals using gut microbiome
data}\label{lack-of-power-and-small-effect-size-confounds-the-ability-to-differentiate-non-obese-and-obese-individuals-using-gut-microbiome-data}

@@

Two recent studies have re-analyzed published data and found that when
datasets are analyzed independently there was limited support the widely
accepted hypothesis that changes in the microbiome are associated with
obesity. This hypothesis was reconsidered by increasing the number of
data sets and pooling the results across the individual datasets. The
Preferred Reporting Items for Systematic Reviews and Meta-Analyses
(PRISMA) guidelines were applied to identify 10 studies for an updated
and more synthetic analysis. Alpha diversity metrics and the relative
risk of obesity based on those metrics were used to identify a limited
number of significant associations with obesity; however, when the
results of the studies were pooled using a random effects model
significant associations were observed between Shannon diversity, number
of observed OTUs, and Shannon evenness and obesity status. They were not
observed for the ratio of \emph{Bacteroidetes} and \emph{Firmicutes} or
their individual relative abundances. Although these tests yielded small
P-values, the difference between the Shannon diversity index of
non-obese and obese individuals was 2.07\%. A power analysis
demonstrated that only one of the studies had sufficient power to detect
a 5\% difference in diversity. When Random Forest machine learning
models were trained on one dataset and then tested using the other 9
datasets, the median accuracy varied between 33.01 and 64.77\%
(median=56.67\%). Although there was support for a relationship between
the microbial communities found in human feces and obesity status, this
association is relatively weak and its detection is confounded by large
interpersonal variation and insufficient sample sizes.

@@

Obesity is a growing health concern with approximately 20\% of the youth
(aged 2-19) in the United States classified as either overweight or
obese (1). This number increases to approximately 35\% in adults (aged
20 or older) and these statistics have seen little change since 2003
(1). Traditionally the body mass index (BMI) has been used as the
traditional method of classifying individuals as non-obese or obese (2).
Recently, there has been increased interest in the role of the
microbiome in modulating obesity (3, 4). If the microbiome does affect
obesity status, then manipulating the microbiome could have a
significant role in the future treatment of obesity and in helping to
stem the current epidemic.

There have been several studies that report observing a link between the
composition of microbiome and obesity in animal models and in humans.
The first such study used genetically obese mice and observed the ratio
of the relative abundances of \emph{Bacteroidetes} to \emph{Firmicutes}
(B:F) was lower in obese mice than lean mice (5). Translation of this
result to humans by the same researchers did not observe this effect,
but did find that obese individuals had a lower diversity than lean
individuals (6). They also showed that the relative abundance of
\emph{Bacteroidetes} and \emph{Firmicutes} increased and decreased,
respectively, as obese individuals lost weight while on a fat or
carbohydrate restricted diet (7). Two re-analysis studies interrogated
previously published microbiome and obesity data and concluded that the
previously reported differences in community diversity and B:F among
non-obese and obese individuals could not be generalized (8, 9).
Regardless of the results using human populations, mechanistic studies
using animal models that were manipulated with antibiotics or
colonization with varied communities appears to support the association
since these manipulation yielded variation in animal weight (10--13).
The purported association between the differences in the microbiome and
obesity have been widely repeated with little attention given to the
lack of a clear signal in human cohort studies.

The recent publication of additional studies that collected BMI data for
each subject as well as other studies that were not included in the
earlier re-analysis studies offered the opportunity to revisit the
question relating the structure of the human microbiome to obesity. One
critique of the prior re-analysis studies is that the authors did not
aggregate the results across studies to increase the effective sample
size. It is possible that there were small associations within each
study that were not statistically significant because the individual
studies lacked sufficient power. Alternatively, diversity metrics may
mask the appropriate signal and it is necessary to measure the
association at the level of microbial populations. Walters et al. (8)
demonstrated that Random Forest machine learning models were capable of
predicting obesity status within a single cohort, but did not attempt to
test the models on other cohorts. The purpose of this study was to
perform a meta-analysis of the association between differences in the
microbiome and obesity status by analyzing and applying a more
systematic and synthetic approach than was used previously.

@@

\textbf{\emph{Literature Review and Study Inclusion.}} We followed the
Preferred Reporting Items for Systematic Reviews and Meta-Analyses
(PRISMA) guidelines to identify studies to include in our meta-analysis
(14). A detailed description of our selection process and the exact
search terms are provided in the Supplemental Text and in Figure 1.
Briefly, we searched PubMed for original research studies that involved
studying obesity and the human microbiome. The initial search yielded
187 studies. We identified number\_string{[}n{]} additional studies that
were not designed to explicitly test for an association between the
microbiome and obesity. We then manually curated the 196 studies to
select those studies that included BMI and sequence data. This yielded
10 eligible studies. An additional study was removed from our analysis
because no individuals in the study had a BMI over 30. Among the final
10 studies, 3 were from identified from our PubMed search (10, 15, 16),
5 were originally identified from the 9 studies that did not explicitly
investigate obesity but included BMI data (17--21), and two datasets
were used (22, 23) because these publications did not specifically look
for any metabolic or obesity conditions but had control populations and
enabled us to help mitigate against publication biases associated with
the bacterial microbiome and obesity. The ten studies are summarized in
Table 1.

\textbf{\emph{Alpha diversity analysis.}} We calculated the Shannon
diversity index, richness, and Shannon evenness, the relative abundance
of \emph{Bacteroidetes} and \emph{Firmicutes}, and the ratio of their
relative abundance (B:F) for each sample. Once we transformed each of
the six alpha diversity metrics to make them normally distributed, we
used a t-test to identify significant associations between the alpha
diversity metric and whether an individual was obese for each of the ten
studies. The B:F and the relative abundance of \emph{Firmicutes} were
not significantly associated with obesity in any study. We identified
seven P-values that were less than 0.05: three studies indicated obese
individuals had a lower richness, two studies indicated a significantly
lower diversity, one study indicated a significantly lower evenness, and
one study indicated a significantly higher relative abundance of
\emph{Bacteroidetes} (Figures 2 and S1). These results largely match
those of the Walters and Finucane re-analysis studies. Interestingly,
although only two of the ten studies observed the previously reported
association between lower diversity and obesity, the other studies
appeared to have the same trend, albeit the differences were not
statistically significant. We used a random effects linear model to
combine the studies using the study as the random effect and found
statistical support for decreased richness, evenness, and diversity
among obese individuals (all P\textless{}0.011). Although there was a
significant relationship between these metrics and obesity status, the
effect size was quite small. The obese individuals averaged 7.47\% lower
richness, 0.88\% lower evenness, and 2.07\% lower diversity. There were
no significant associations when we pooled the phylum-level metrics
across studies. These results indicate that obese individuals do have a
statistically significant lower diversity than non-obese individuals;
however, it is questionable whether the difference is biologically
significant.

\textbf{\emph{Relative risk.}} Building upon the alpha diversity
analysis we calculated the relative risk of being obese based on an
individual's alpha diversity metrics relative to the median metric for
that study. The results using relative risk largely matched those of
using the untransformed alpha diversity data. Across the ten studies and
six metrics, the only significant relative risk values were the
richness, evenness, and diversity values from the Goodrich study
(Figures 3 and S2). Again, although the relative risk values were not
significant for other studies, the values tended to be above one. When
we pooled the data using a random effects model, the relative risk
associated with having a richness, evenness, or diversity below the
median for the population was significantly associated with obesity (all
P\textless{}0.0044). The relative risks associated with alpha diversity
were small. The relative risk of having a low richness was 1.30
(1.13-1.49), low evenness was 1.20 (1.06-1.37), and low diversity was
1.27 (1.09-1.48). There were no significant difference in the
phylum-level metrics. Again, the relative risk results indicate that
individuals with a lower richness, evenness, or diversity are at
statistically significant increased risk of being obese, it is
questionable whether that risk is biologically or clinically relevant.

\textbf{\emph{Beta diversity analysis.}} Following the approach used by
the Walters and Finucane re-analysis studies, for each dataset we
calculated a Bray-Curtis distance matrix to measure the difference in
the membership and structure of the individuals from each study. We then
used AMOVA to test for a significant differences between the structure
of non-obese and obese individuals (Table 1). The Escobar, Goodrich, and
Turnbaugh datasets indicated a significant difference in community
structure (all P\textless{}0.05). Because it was not possible to
ascertain the directionality of the difference in community structure
nor perform a pooled analysis using studies that had non-overlapping 16S
rRNA gene sequence regions it is unclear whether these differences
reflect a broader, but perhaps small, shift in community structure
between non-obese and obese individuals.

\textbf{\emph{Development of a microbiome-based classifier of obesity.}}
The Walters re-analysis study suggested that it was possible to classify
individuals as being non-obese or obese based on the composition of
their microbiota. We repeated this analysis with additional datasets
using OTU and genus-level phylotype data. For each study we developed a
Random Forest machine learning model to classify individuals. Using
ten-fold cross validation, the observed AUC values varied between 0.52
and 0.69 indicating a relatively poor ability to classify individuals
(Figure 4A). So that we could test models on other datasets, we trained
models using genus-level phylotype data for each dataset. The
cross-validated AUC values for the models applied to the training
datasets varied between 0.51 and 0.65, again indicating a relatively
poor ability to classify individuals from the original dataset (Figure
4B). For each model we identified the probability where the sum of the
sensitivity and specificity was the highest. We then used this
probability to define a threshold for calculating the accuracy of the
models when applied to the other nine datasets (Figure 5). Although
there considerable variation in accuracy values for each model, the
median accuracy for each model varied between 0.33 (Turnbaugh) and 0.65
(HMP) (median=0.57). When we considered the number of samples, balance
of non-obese and obese individuals, and region within the 16S rRNA gene
it was not possible to identify factors that predictably affected model
performance. The ability to predict obesity status using the relative
abundance of OTUs and genera in the communities is only marginally
better than random. These results suggest that given the large diversity
of microbiome compositions it is difficult to identify a taxonomic
signal that can be associated with obesity.

\textbf{\emph{Power and Sample Size Estimate Simulations.}} The
inability to detect a difference between non-obese and obese individuals
could be due to the lack of a true effect or because the study had
insufficient statistical power to detect a difference because of
insufficient sampling, large interpersonal variation, and unbalanced
sampling of non-obese and obese individuals. To assess this, we
calculated the power to detect differences of 1, 5, 10, and 15\% in each
of the alpha diversity metrics using the sample sizes used in each of
the studies (Figures 6, S3-S8). Although there is no biological
rationale for these effect sizes, they represent a range that is
plausible. Only the Goodrich study had power greater than 0.80 to detect
a 5\% difference in Shannon diversity and six of the studies had enough
power to detect a 10\% difference (Figure 6). None of the studies had
sufficient power to detect a 15\% difference between B:F values (Figure
S5). In fact, the maximum power among any of the studies to detect a
15\% difference in B:F values was 0.25. Among the tests for relative
risk, none of the studies had sufficient power to detect a Cohen's d of
0.10 and only two studies had sufficient power to detect a Cohen's d of
0.15. We next estimated how many individuals would need to have been
sampled to have sufficient power to detect the four effect sizes
assuming the observed interpersonal variation from each study and
balanced sampling between the two groups. To detect a 1, 5, 10, or 15\%
difference in Shannon index, the median required sampling effort per
group was approximately 3,400, 140, 35, or 16 individuals, respectively.
To detect a 1, 5, 10, and 15\% difference in B:F values, the median
required sampling effort per group was approximately 160,000, 6,300,
1,600, or 700 individuals, respectively. To detect a 1, 5, 10, and 15\%
difference in relative risk values using Shannon diversity, the median
required sampling effort per group was approximately 39,000, 1,500, 380,
or 170 individuals, respectively. These estimates indicate that most
microbiome studies are underpowered to detect modest effect sizes using
either metric. In the case of obesity, the studies were underpowered to
detect the \texttt{signif(range\_effect\_size,\ 1){[}1{]}} to 6\%
difference in diversity that was observed across the studies.

@@

Our meta-analysis helps to provide a clarity to the ongoing debate of
whether or not there are specific microbiome-based markers that can be
associated with obesity. We performed an extensive literature review of
the existing studies on the microbiome and obesity and performed a
meta-analysis on the studies that remained based on our inclusion and
exclusion criteria. By statistically pooling the data from ten studies,
we observed significant, but small, relationships between richness,
evenness, and diversity and obesity status as well as the relative risk
of being obese based on these metrics. We also generated Random Forest
machine learning models trained on each dataset and tested on the
remaining datasets. This analysis demonstrated that the ability to
reliably classify individuals as being obese based on the composition of
their microbiome was limited. Finally, we assessed the ability of each
study to detect defined differences in alpha diversity and observed that
most studies were underpowered to detect modest effect sizes.
Considering these datasets are among the largest published, it appears
that most microbiome studies are underpowered to detect differences in
alpha diversity.

@@

By selecting a range of possible effect sizes, we were able to
demonstrate that most studies are underpowered to detect modest
differences in alpha diversity metrics and phylum-level relative
abundances. Several factors interact to limit the power of microbiome
studies. There is wide interpersonal variation in the diversity and
structure of the human microbiome. In addition, the common experimental
designs limit their power. As we observed, most of the studies included
in our analysis were unbalanced for the variable that we were interested
in. This was also true of those studies that originally sought to
identify associations with obesity. Even with a balanced design, we
showed that it was necessary to obtain approximately 140 and 6,300
sequences per sample to detect a 5\% difference in Shannon diversity or
B:F, respectively. It was interesting that these sample sizes agreed
across studies regardless of their sequencing method, region with in the
16S rRNA gene, or subject population (Figure 6). This suggests that
regardless of the treatment or category, these sample sizes represent a
good starting point for subject recruitment when using stool samples.
Unfortunately, few studies have been published with this level of
subject recruitment. This is troubling since the positive predictive
rate of a significant finding in an underpowered study is small leading
to results that cannot be reproduced (24). Future microbiome studies
should articulate the basis for their experimental design.

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

\section{Looking for a Signal in the Noise: Revisiting Obesity and the
Microbiome}\label{looking-for-a-signal-in-the-noise-revisiting-obesity-and-the-microbiome}

@@

Two recent studies have re-analyzed published data and found that when
datasets are analyzed independently there was limited support for the
widely accepted hypothesis that changes in the microbiome are associated
with obesity. This hypothesis was reconsidered by increasing the number
of datasets and pooling the results across the individual datasets. The
Preferred Reporting Items for Systematic Reviews and Meta-Analyses
(PRISMA) guidelines were applied to identify 10 studies for an updated
and more synthetic analysis. Alpha diversity metrics and the relative
risk of obesity based on those metrics were used to identify a limited
number of significant associations with obesity; however, when the
results of the studies were pooled using a random effects model
significant associations were observed between Shannon diversity, number
of observed OTUs, and Shannon evenness and obesity status. They were not
observed for the ratio of \emph{Bacteroidetes} and \emph{Firmicutes} or
their individual relative abundances. Although these tests yielded small
P-values, the difference between the Shannon diversity index of
non-obese and obese individuals was 2.07\%. A power analysis
demonstrated that only one of the studies had sufficient power to detect
a 5\% difference in diversity. When Random Forest machine learning
models were trained on one dataset and then tested using the other 9
datasets, the median accuracy varied between 33.01 and 64.77\%
(median=56.68\%). Although there was support for a relationship between
the microbial communities found in human feces and obesity status, this
association is relatively weak and its detection is confounded by large
interpersonal variation and insufficient sample sizes.

@@

Obesity is a growing health concern with approximately 20\% of the youth
(aged 2-19) in the United States classified as either overweight or
obese (1). This number increases to approximately 35\% in adults (aged
20 or older) and these statistics have seen little change since 2003
(1). Traditionally, the body mass index (BMI) has been used to classify
individuals as non-obese or obese (2). Recently, there has been
increased interest in the role of the microbiome in modulating obesity
(3, 4). If the microbiome does affect obesity status, then manipulating
the microbiome could have a significant role in the future treatment of
obesity and in helping to stem the current epidemic.

There have been several studies that report observing a link between the
composition of microbiome and obesity in animal models and in humans.
The first such study used genetically obese mice and observed the ratio
of the relative abundances of \emph{Bacteroidetes} to \emph{Firmicutes}
(B:F) was lower in obese mice than lean mice (5). Translation of this
result to humans by the same researchers did not observe this effect,
but did find that obese individuals had a lower alpha-diversity than
lean individuals (6). They also showed that the relative abundance of
\emph{Bacteroidetes} and \emph{Firmicutes} increased and decreased,
respectively, as obese individuals lost weight while on a fat or
carbohydrate restricted diet (7). Two re-analysis studies by Walters et
al. (8) and Finucane et al. (9) interrogated previously published
microbiome and obesity data and concluded that the previously reported
differences in community diversity and B:F among non-obese and obese
individuals could not be generalized. Regardless of the results using
human populations, studies using animal models where the community was
manipulated with antibiotics or established by colonizing germ-free
animals with varied communities appear to support the association since
these manipulations yielded differences in animal weight (10--13). The
purported association between the differences in the microbiome and
obesity have been widely repeated with little attention given to the
lack of a clear signal in human cohort studies.

The recent publication of additional studies that collected BMI data for
each subject as well as other studies that were not included in the
earlier re-analysis studies offered the opportunity to revisit the
question relating the structure of the human microbiome to obesity. One
critique of the prior re-analysis studies is that the authors did not
aggregate the results across studies to increase the effective sample
size. It is possible that there were small associations within each
study that were not statistically significant because the individual
studies lacked sufficient power. Alternatively, diversity metrics may
mask the appropriate signal and it is necessary to measure the
association at the level of microbial populations. The Walters
re-analysis study demonstrated that Random Forest machine learning
models were capable of predicting obesity status within a single cohort,
but did not attempt to test the models on other cohorts. The purpose of
this study was to perform a meta-analysis of the association between
differences in the microbiome and obesity status by analyzing and
applying a more systematic and synthetic approach than was used
previously.

@@

\textbf{\emph{Literature Review and Study Inclusion.}} To perform a
robust meta-analysis and limit inclusion bias, we followed the Preferred
Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA)
guidelines to identify the studies that we analyzed (14). A detailed
description of our selection process and the exact search terms are
provided in the Supplemental Text and in Figure 1. Briefly, we searched
PubMed for original research studies that involved studying obesity and
the human microbiome. The initial search yielded 187 studies. We
identified ten additional studies that were not designed to explicitly
test for an association between the microbiome and obesity. We then
manually curated the 196 studies to select those studies that included
BMI and 16S rRNA gene sequence data. This yielded 10 eligible studies.
An additional study was removed from our analysis because no individuals
in the study had a BMI over 30. Among the final 10 studies, 3 were from
identified from our PubMed search (10, 15, 16), 5 were originally
identified from the 10 studies that did not explicitly investigate
obesity but included BMI data (17--21), and two datasets were used (22,
23) because these publications did not specifically look for any
metabolic or obesity conditions but had control populations and enabled
us to help mitigate against publication biases associated with the
bacterial microbiome and obesity. The ten studies are summarized in
Table 1. For comparison, two of these studies were included in the
Finucane re-analysis study (10, 21) and four of these studies were
included in the Walters re-analysis study (10, 15, 20, 21). The 16S rRNA
gene sequence data from each study was re-analyzed using a similar
approach based on previously described methods for reducing the number
of chimeric sequences and sequencing errors for 454 and Illumina MiSeq
data (24, 25). The sequences were either clustered into operational
taxonomic units (OTUs) using the average neighbor approach (26) or
taxonomic groupings based on their classification using a naive Bayesian
classifier (27).

\textbf{\emph{Alpha diversity analysis.}} We calculated the Shannon
diversity index, observed richness, and Shannon evenness, the relative
abundance of \emph{Bacteroidetes} and \emph{Firmicutes}, and the ratio
of their relative abundance (B:F) for each sample. Once we transformed
each of the six alpha diversity metrics to make them normally
distributed, we used a t-test to identify significant associations
between the alpha diversity metric and whether an individual was obese
for each of the ten studies. The B:F and the relative abundance of
\emph{Firmicutes} were not significantly associated with obesity in any
study. We identified seven P-values that were less than 0.05: three
studies indicated obese individuals had a lower richness, two studies
indicated a significantly lower diversity, one study indicated a
significantly lower evenness, and one study indicated a significantly
higher relative abundance of \emph{Bacteroidetes} (Figures 2 and S1).
These results largely match those of the Walters and Finucane
re-analysis studies. Interestingly, although only two of the ten studies
observed the previously reported association between lower diversity and
obesity, the other studies appeared to have the same trend, albeit the
differences were not statistically significant. We used a random effects
linear model to combine the studies using the study as the random effect
and found statistical support for decreased richness, evenness, and
diversity among obese individuals (all P\textless{}0.011). Although
there was a significant relationship between these metrics and obesity
status, the effect size was quite small. The obese individuals averaged
7.47\% lower richness, 0.88\% lower evenness, and 2.07\% lower
diversity. There were no significant associations when we pooled the
phylum-level metrics across studies. These results indicate that obese
individuals do have a statistically significant lower diversity than
non-obese individuals; however, it is questionable whether the
difference is biologically significant.

\textbf{\emph{Relative risk.}} Building upon the alpha diversity
analysis we calculated the relative risk of being obese based on an
individual's alpha diversity metrics relative to the median metric for
that study. Inspection of funnel plots for each of the metrics suggested
that the studies included in our analysis were not biased (Figure S2).
The results using relative risk largely matched those of using the raw
alpha diversity data. Across the ten studies and six metrics, the only
significant relative risk values were the richness, evenness, and
diversity values from the Goodrich study (Figures 3 and S3). Again,
although the relative risk values were not significant for other
studies, the values tended to be above one. When we pooled the data
using a random effects model, the relative risk associated with having a
richness, evenness, or diversity below the median for the population was
significantly associated with obesity (all P\textless{}0.0044). The
relative risks associated with alpha diversity were small. The relative
risk of having a low richness was 1.30 (95\% CI: 1.13-1.49), low
evenness was 1.20 (95\% CI: 1.06-1.37), and low diversity was 1.27 (95\%
CI: 1.09-1.48). There were no significant differences in the
phylum-level metrics. Again, the relative risk results indicate that
individuals with a lower richness, evenness, or diversity are at
statistically significant increased risk of being obese, it is
questionable whether that risk is biologically or clinically relevant.

\textbf{\emph{Beta diversity analysis.}} Following the approach used by
the Walters and Finucane re-analysis studies, for each dataset we
calculated a Bray-Curtis distance matrix to measure the difference in
the membership and structure of the individuals from each study. We then
used AMOVA to test for significant differences between the structure of
non-obese and obese individuals (Table 1). The Escobar, Goodrich, and
Turnbaugh datasets indicated a significant difference in community
structure (all P\textless{}0.05). Because it was not possible to
ascertain the directionality of the difference in community structure
because the samples are arrayed in a non-dimensional space. nor perform
a pooled analysis using studies that had non-overlapping 16S rRNA gene
sequence regions it is unclear whether these differences reflect a
broader, but perhaps small, shift in community structure between
non-obese and obese individuals.

\textbf{\emph{Development of a microbiome-based classifier of obesity.}}
The Walters re-analysis study suggested that it was possible to classify
individuals as being non-obese or obese based on the composition of
their microbiota. We repeated this analysis with additional datasets
using OTU and genus-level phylotype data. For each study we developed a
Random Forest machine learning model to classify individuals. Using
ten-fold cross validation, the observed AUC values varied between 0.52
and 0.69 indicating a relatively poor ability to classify individuals
(Figure 4A). So that we could test models on other datasets, we trained
models using genus-level phylotype data for each dataset. The
cross-validated AUC values for the models applied to the training
datasets varied between 0.51 and 0.65, again indicating a relatively
poor ability to classify individuals from the original dataset (Figure
4B). For each model we identified the probability where the sum of the
sensitivity and specificity was the highest. We then used this
probability to define a threshold for calculating the accuracy of the
models when applied to the other nine datasets (Figure 5). Although
there was considerable variation in accuracy values for each model, the
median accuracy for each model varied between 0.33 (Turnbaugh) and 0.65
(HMP) (median=0.57). When we considered the number of samples, balance
of non-obese and obese individuals, and region within the 16S rRNA gene
it was not possible to identify factors that predictably affected model
performance. The ability to predict obesity status using the relative
abundance of OTUs and genera in the communities is only marginally
better than random. These results suggest that given the large diversity
of microbiome compositions it is difficult to identify a taxonomic
signal that can be associated with obesity.

\textbf{\emph{Power and Sample Size Estimate Simulations.}} The
inability to detect a difference between non-obese and obese individuals
could be due to the lack of a true effect or because the study had
insufficient statistical power to detect a difference because of
insufficient sampling, large interpersonal variation, and unbalanced
sampling of non-obese and obese individuals. To assess this, we
calculated the power to detect differences of 1, 5, 10, and 15\% in each
of the alpha diversity metrics using the sample sizes used in each of
the studies (Figures 6, S4-S9). Although there is no biological
rationale for these effect sizes, they represent a range that includes
effect sizes that would be generally considered to be biologically
significant. Only the Goodrich study had power greater than 0.80 to
detect a 5\% difference in Shannon diversity and six of the studies had
enough power to detect a 10\% difference (Figure 6). None of the studies
had sufficient power to detect a 15\% difference between B:F values
(Figure S4). In fact, the maximum power among any of the studies to
detect a 15\% difference in B:F values was 0.25. Among the tests for
relative risk, none of the studies had sufficient power to detect a
Cohen's d of 0.10 and only two studies had sufficient power to detect a
Cohen's d of 0.15. We next estimated how many individuals would need to
have been sampled to have sufficient power to detect the four effect
sizes assuming the observed interpersonal variation from each study and
balanced sampling between the two groups. To detect a 1, 5, 10, or 15\%
difference in Shannon index, the median required sampling effort per
group was approximately 3,400, 140, 35, or 16 individuals, respectively.
To detect a 1, 5, 10, and 15\% difference in B:F values, the median
required sampling effort per group was approximately 160,000, 6,300,
1,600, or 700 individuals, respectively. To detect a 1, 5, 10, and 15\%
difference in relative risk values using Shannon diversity, the median
required sampling effort per group was approximately 39,000, 1,500, 380,
or 170 individuals, respectively. These estimates indicate that most
microbiome studies are underpowered to detect modest effect sizes using
either metric. In the case of obesity, the studies were underpowered to
detect the 0.90 to 6\% difference in diversity that was observed across
the studies.

@@

Our meta-analysis helps to provide clarity to the ongoing debate of
whether or not there are specific microbiome-based markers that can be
associated with obesity. We performed an extensive literature review of
the existing studies on the microbiome and obesity and performed a
meta-analysis on the studies that remained based on our inclusion and
exclusion criteria. By statistically pooling the data from ten studies,
we observed significant, but small, relationships between richness,
evenness, and diversity and obesity status as well as the relative risk
of being obese based on these metrics. We also generated Random Forest
machine learning models trained on each dataset and tested on the
remaining datasets. This analysis demonstrated that the ability to
reliably classify individuals as being obese based solely on the
composition of their microbiome was limited. Finally, we assessed the
ability of each study to detect defined differences in alpha diversity
and observed that most studies were underpowered to detect modest effect
sizes. Considering these datasets are among the largest published, it
appears that most human microbiome studies are underpowered to detect
differences in alpha diversity.

@@

By selecting a range of possible effect sizes, we were able to
demonstrate that most studies are underpowered to detect modest
differences in alpha diversity metrics and phylum-level relative
abundances. Several factors interact to limit the power of microbiome
studies. There is wide interpersonal variation in the diversity and
structure of the human microbiome. Some factors such as relationship
between subjects could potentially decrease the amount variation (6) and
other factors such as whether one lives in a rural environment could
increase the amount of variation (28). It was interesting that although
there were ranges in variation across the studies included in our
analysis, the sample size calculations yielded similar results. In
addition, the common experimental designs limit their power. As we
observed, most of the studies included in our analysis were unbalanced
for the variable that we were interested in. This was also true of those
studies that originally sought to identify associations with obesity.
Even with a balanced design, we showed that it was necessary to obtain
approximately 140 and 6,300 samples per group to detect a 5\% difference
in Shannon diversity or B:F, respectively. It was interesting that these
sample sizes agreed across studies regardless of their sequencing
method, region within the 16S rRNA gene, or subject population (Figure
6). This suggests that regardless of the treatment or category, these
sample sizes represent a good starting point for subject recruitment
when using stool samples. Unfortunately, few studies have been published
with this level of subject recruitment. This is troubling since the
positive predictive rate of a significant finding in an underpowered
study is small leading to results that cannot be reproduced (29). Future
microbiome studies should articulate the basis for their experimental
design.
